From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-30 Pi+/paper/equivalence.tex
Outlining more in section 7

\begin{toappendix}
  \todo{What is U?}
  \begin{proposition}
    We can form  weak 2-category $\PiCat$ with
    \begin{itemize}
      \item $\PiLang$ types for 0-cells,
      \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
      \item for $p, q : X \iso Y$, a collection of 2-cells $p \Iso q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form  weak 2-category $\PiPlusCat$ with
    \begin{itemize}
      \item $\PiPlusLang$ types for 0-cells,
      \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
      \item for $p, q : X \isop Y$, a collection of 2-cells $p \Isop q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form  weak 2-category $\PiHatCat$ with
    \begin{itemize}
      \item $\PiHatCat$ types for 0-cells,
      \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
      \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Isoh q$.
    \end{itemize}
  \end{proposition}
\end{toappendix}

@@

Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
$\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
adjacent transpositions, and an appropriate set of 2-combinators. The syntax, again omitting 2-combinators, is given
in~\cref{fig:pihat}. \vc{Something about PROPs.}

We normalise $\PiPlusLang$ to $\PiHatLang$ and back, by giving $\evalh/\quoteh$ maps for 0, 1, and 2-cells.

@@

@@

\vc{Define maps quote/eval on 0,,2-cells, and show commuting diagrams.}

\begin{proposition}
  We can form a weak 2-category $\PiHatCat$ with
  \begin{itemize}
    \item natural numbers for 0-cells,
    \item for $n, m : \Nat$, a collection of 1-cells $n \isoh m$,
    \item for $p, q : n \isoh m$, a collection of 2-cells $p \Isoh q$.
  \end{itemize}
\end{proposition}

\begin{proposition}
  There is a symmetric monoidal structure on $\PiHatCat$, with $0$ for the unit, and addition for the tensor.
\end{proposition}

Then we establish the completeness of $\PiHatLang$ with respect to $\UFin$. We define $\evalt$ and $\quotet$ for 0, 1,
and 2-cells.

@@

@@

@@

To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote back it to
$\PiPlusLang$.

>>>>

\begin{toappendix}
  \todo{What is U?}
  \begin{proposition}
    We can form a weak 2-category $\PiCat$ with
    \begin{itemize}
      \item $\PiLang$ types for 0-cells,
      \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
      \item for $p, q : X \iso Y$, a collection of 2-cells $p \Iso q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form a weak 2-category $\PiPlusCat$ with
    \begin{itemize}
      \item $\PiPlusLang$ types for 0-cells,
      \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
      \item for $p, q : X \isop Y$, a collection of 2-cells $p \Isop q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form a weak 2-category $\PiHatCat$ with
    \begin{itemize}
      \item $\PiHatCat$ types for 0-cells,
      \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
      \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Isoh q$.
    \end{itemize}
  \end{proposition}
\end{toappendix}

@@

Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
$\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
adjacent transpositions, and an appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators,
in~\cref{fig:pihat}. \vc{Something about PROPs.}

As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers.
However, we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture.

\begin{proposition}
  $\PiHatCat$ has a symmetric monoidal structure, with the unit given by 0 and the tensor product given by natural
  number addition.
\end{proposition}

Using this symmetric monoidal structure, we translate from $\PiPlusLang$ to $\PiHatLang$.

@@

To go back from $\PiHatLang$ to $\PiPlusLang$, we turn a natural number into a $\PiPlusLang$ type, using right-biased
addition, that is, the natural number $n$ gets mapped to the type $\onet + (\onet + (\onet + \ldots + \zerot))$. The
non-trivial part is coherently generating a $\PiPlusLang$ combinator from $\PiHatLang$ combinators, using adjacent swaps
to generate big swaps.

@@

Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and go back from $\UFin$ to $\PiHatLang$.

\todo{Sn/Lehmer/Aut(Fin n) diagram here?}

Types in $\PiHatLang$ are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The
1-combinators in $\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, or permutations, in
each connected component. Finally, 2-combinators are mapped to 2-paths between loops in $\UFin$.

@@

0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to generate
a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. 2-paths are quoted back to
2-combinators in $\PiHatLang$.

@@

$\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category.

@@

To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote it back to
$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ to $\PiHatLang$.
